---
title: "Volleyball Percentiles"
---

We'll be calculating percentiles to see how good Maryland was in service aces in the 2024 season. We'll use `mutate()` and a new function `ntile()` to calculate percentiles. You should open the VBPercentiles_Explainer.Rmd document in the explainers folder and use that as a reference.

Get started by loading the tidyverse in a code block, then load the CSV file here: https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv" and save it to a dataframe called `matches`.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the data

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

### Task 3: Create totals of aces for each team using `group_by()` and `summarize()`

```{r}
team_aces <- matches |> 
  group_by(team) |> 
  summarize(total_aces = sum(aces)) |> 
  arrange(desc(total_aces))
```


### Task 4: Calculate percentiles using `mutate()` and `ntile()`

```{r}
team_aces |> mutate(percentile = ntile(total_aces, 100))
```


### Task 5: Describe the percentiles

Write a couple of sentences that puts Maryland into context with the other teams in the dataset. You can do some Internet research to help with this. Maryland is in the 95th percentile with 223 aces among the top 17 schools in womens volleyball. Yet when I look at their record online they had an overall record of 14-17. Conference Record: 5-15 (.250) Home Record: 8-7 Away Record: 4-9. Utep meanwhile had an overall Record of 27–6, Home Record: 15–3Away Record: 8–2Neutral Record: 4–1Postseason: Earned the program's first-ever at-large bid to the NCAA Tournament.Milestones: Set a school-record 13-match win streak and became the first Division I team in the country to reach 20 wins during the season. Seems Maryland was in a tough division that year.



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.






