---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE: The report lists 411 people in three employment categories: 365 employees, 43 detailees, and three part-time detailees. From a reporting standpoint, the repeated or vague job titles, including 21 senior policy advisors, stand out, while the 10 salaries recorded as $0 could lower the mean and require further reporting to determine whether those individuals are unpaid or compensated through another agency.


### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
  summarize(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY)
  )

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
  summarize(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY),
    min_salary = min(SALARY),
    max_salary = max(SALARY)
  )

```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE: The mean salary is about $112,703 and the median is $109,243, showing that salaries are centered near $110,000, with higher-paid positions raising the average slightly. Salaries range from $0 to $197,200; the 10 zero-dollar salaries are unusual records that require explanation, while the many salaries near the maximum form a high-pay cluster rather than isolated outliers.


### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE: My best story idea is to examine why the nine highest‑paid people in the 2026 White House salary report are all detailees, including three listed as part‑time. This is a specific and verifiable pattern in the data, and it raises clear questions about how these assignments work and which agencies actually fund their salaries. My next step would be to identify each detailee’s home agency and request their detail agreements, which would show their duties, working arrangements, and whether the White House or the originating agency pays their salary.
